Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-10-02 submission/NatCom_methods.tex
more on address presubmission comments

Recall that we can express the objective/optimization function for a typical DNN with $L$ layers, with activation functions $h_{l}(\cdot)$, and with $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$ as Equation~(\ref{eqn:dnn_energy}).
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.
Typically, this model would be trained on some labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using Backprop, by minimizing the loss $\mathcal{L}$.
For simplicity, we do not indicate the structural details of the layers (e.g., Dense or not, Convolutions or not, Residual/Skip Connections, etc.). 
Each layer is defined by one or more layer 2D weight matrices $\mathbf{W}_{l}$, and/or the 2D feature maps $\mathbf{W}_{l,i}$ extracted from 2D Convolutional (Conv2D) layers.
A typical modern DNN may have anywhere between 5 and 5000 2D layer matrices.

>>>>

Recall that we can express the objective/optimization function for a typical DNN with $L$ layers and with $N\times M$ weight matrices $\mathbf{W}_{l}$ and bias vectors $\mathbf{b}_{l}$ as Equation~(\ref{eqn:dnn_energy}).
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.
Typically, this model would be trained on some labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using Backprop, by minimizing the loss $\mathcal{L}$.
For simplicity, we do not indicate the structural details of the layers (e.g., Dense or not, Convolutions or not, Residual/Skip Connections, etc.). 
Each layer is defined by one or more layer 2D weight matrices $\mathbf{W}_{l}$, and/or the 2D feature maps $\mathbf{W}_{l,i}$ extracted from 2D Convolutional (Conv2D) layers.
A typical modern DNN may have anywhere between 5 and 5000 2D layer~matrices.

### 2020-10-02 submission/NatCom_overall_approach.tex
adjust img2 to img
more on address presubmission comments

Consider the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, activation functions $h_{l}(\cdot)$, and $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$, as 
the minimization of a general Loss Function $\mathcal{L}$ over $N$ training data instances and labels, $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$.
For a typical supervised classification problem, one has
\begin{equation}
\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N}\left(\mathcal{L}(E_{DNN}(\mathbf{x}_{i})-y_{i})\right)  ,
\end{equation}
where the loss function $\mathcal{L}(\cdot)$ can take on a myriad of forms~\cite{JC17_TR}, and where
\begin{equation}
E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})
\label{eqn:dnn_energy}
\end{equation}
is the Energy (or optimization) Landscape function.  
\michael{See comments for math equations, explain $h_{L}$, scalar valued function, $b_L$, times symbol, etc., confusing about role or $y$, and whether we are dealing with products of matrices.  E.g., $L$ layers, with activation functions $h_{l}(\cdot)$, and with $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$; etc.  }
$E_{DNN}$ does not explicitly depend on the data, but rather it maps data instance vectors ($\mathbf{x}_i$ values) to predictions (labels).
\michael{Mark found that sentence confusing.}
Therefore, one can analyze $E_{DNN}$ in the absence of any training or test~data.

@@

Our approach involves analyzing individual DNN weight matrices, for (depending on the architecture) fully-connected and/or convolutional layers.
Each DNN layer contains one or more layer 2D  $N\times M$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, e.g., extracted from 2D Convolutional layers, where $N > M$.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
(We may drop the $i$ and/or $i,l$ subscripts below.)
The best performing quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's Empirical Correlation Matrix, $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
We consider the following metrics:
\begin{eqnarray}
& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
& & \text{$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
\end{eqnarray}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
These eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, $\lambda_{i}=\sigma^{2}_{i}$.
All four metrics can be computed easily from DNN weight matrices.
The first two metrics are well-known in ML.
The last two metrics deserve special mention, as they depend on an empirical parameter $\alpha$ that is the PL exponent that arises in the recently-developed Heavy Tailed Self Regularization (HT-SR) Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.

@@

The original work on HT-SR Theory considered a small number of NNs, including AlexNet and InceptionV3. 
It showed that for nearly every $\mathbf{W}$, the (bulk and tail) of the ESDs can be fit to a truncated PL, and that PL exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
As for the mechanism responsible for these properties, statistical physics offers several possibilities~\cite{SornetteBook,nishimori01}, e.g., self organized criticality~\cite{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\cite{HodMah20A_TR,SorCon97}.
Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR Theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models.
Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

@@

\begin{figure}[t]
    \centering
    %\includegraphics[width=15.0cm]{img/WeightWatcher_v2}
    \includegraphics[width=15.0cm]{../img2/WeightWatcher_v2}
    \caption{Schematic of analyzing DNN layer weight matrices $\mathbf{W}$.  
             Given an individual layer weight matrix $\mathbf{W}$, from either a fully-connected layer or a convolutional layer, perform a Singular Value Decomposition (SVD) to obtain $\mathbf{W} = \mathbf{U} \mathbf{\Sigma} \mathbf{V}^{T}$, and examine the histogram of eigenvalues of $\mathbf{W}^{T}\mathbf{W}$.
             Norm-based metrics (not shown) and PL-based metrics (that depend on fitting the histogram of eigenvalues to a truncated PL) can be used to compare models.
             For example, one can analyze one layer of a pre-trained model, compare multiple layers of a pre-trained model, make comparisons across model architectures, monitor neural network properties during training, etc. 
             \michael{Put the Analyze, Compare, Monitor comments in the caption, not the figure.}
}
    \label{fig:ww}
\end{figure}

>>>>

Consider the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, and weight matrices $\mathbf{W}_{l}$ and bias vectors $\mathbf{b}_{l}$, as 
the minimization of a general Loss Function $\mathcal{L}$ over the training data instances and labels, $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$.
For a typical supervised classification problem, the goal of training is to construct $\mathbf{W}_{l}$ and $\mathbf{b}_{l}$ that capture correlations in the data, in the sense of solving
\begin{equation}
\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N}\left(\mathcal{L}(E_{DNN}(\mathbf{x}_{i})-y_{i})\right)  ,
\end{equation}
where the loss function $\mathcal{L}(\cdot)$ can take on a myriad of forms~\cite{JC17_TR}, and where the Energy (or optimization) Landscape function
\begin{equation}
%E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})
E_{DNN} = f(\mathbf{W}_{1},\ldots,\mathbf{W}_{L},\mathbf{b}_{1},\ldots,\mathbf{b}_{L})
\label{eqn:dnn_energy}
\end{equation}
depends on the weights and biases.
\michael{See comments for math equations.}
For a trained model, $E_{DNN}$ does not explicitly depend on the data, but rather it maps data instance vectors ($\mathbf{x}_i$ values) to predictions ($y_{i}$ labels).
\michael{Mark found that sentence confusing.}
Therefore, one can analyze $E_{DNN}$ in the absence of any training or test~data.

@@

Our approach involves analyzing individual DNN weight matrices, for (depending on the architecture) fully-connected and/or convolutional layers.
Each DNN layer contains one or more layer 2D  $N_{l}\times M_{l}$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, e.g., extracted from 2D Convolutional layers, where $N > M$.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
(We may drop the $i$ and/or $i,l$ subscripts below.)
The best performing quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's Empirical Correlation Matrix, $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
We consider the following metrics:
\begin{eqnarray}
& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
& & \text{$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
\end{eqnarray}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
These eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, $\lambda_{i}=\sigma^{2}_{i}$.
All four metrics can be computed easily from DNN weight matrices.
The first two metrics are well-known in ML.
The last two metrics deserve special mention, as they depend on an empirical parameter $\alpha$ that is the PL exponent that arises in the recently-developed Heavy Tailed Self Regularization (HT-SR) Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.

@@

The original work on HT-SR Theory considered a small number of NNs, including AlexNet and InceptionV3. 
It showed that for nearly every $\mathbf{W}$, the (bulk and tail) of the ESDs can be fit to a truncated PL, and that PL exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
As for the mechanism responsible for these properties, statistical physics offers several possibilities~\cite{SornetteBook,nishimori01}, e.g., self organized criticality~\cite{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\cite{HodMah20A_TR,SorCon97}.
Alternatively, related techniques have been used to analyze correlations and information propogation in actual spiking neurons~\cite{SYYRP11,YKYP14}.
Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR Theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models.
Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

@@

\begin{figure}[t]
    \centering
    %\includegraphics[width=15.0cm]{img/WeightWatcher_v2}
    \includegraphics[width=15.0cm]{img/WeightWatcher_v3}
    \caption{Schematic of analyzing DNN layer weight matrices $\mathbf{W}$.  
             Given an individual layer weight matrix $\mathbf{W}$, from either a fully-connected layer or a convolutional layer, perform a Singular Value Decomposition (SVD) to obtain $\mathbf{W} = \mathbf{U} \mathbf{\Sigma} \mathbf{V}^{T}$, and examine the histogram of eigenvalues of $\mathbf{W}^{T}\mathbf{W}$.
             Norm-based metrics (not shown) and PL-based metrics (that depend on fitting the histogram of eigenvalues to a truncated PL) can be used to compare models.
             For example, one can analyze one layer of a pre-trained model, compare multiple layers of a pre-trained model, make comparisons across model architectures, monitor neural network properties during training, etc. 
}
    \label{fig:ww}
\end{figure}
